Statistical Inference for inter-arrival times of extreme events in bursty time series

### 2020-03-01 article_springer/article_springer.Rmd
Further changes

*Proof of Theorem:* We interpret the threshold crossing time $T(\ell)$ as the hitting time of the 
underlying CTRM (Continuous Time Random Maxima) or "max-renewal process", 
and then utilize a result by @MeerschaertStoev08. 
The running maximum process is defined as 
$$
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
$$
and since we assume that the $J_k$ have a continuous distribution, there exist 
norming functions $a(c)$ and $d(c)$ such that 
$$
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad t \to \infty
$$
where $F$ is a generalized extreme value distribution, and $\ell^*$ is any 
value from the support of $F$. 
The CTRM process is then defined via 
$$
V(t) = M(N(t)), \quad t \ge 0
$$
where $N(t)$ is the renewal process associated with the waiting times 
$W_k$: 
$$
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
$$
Now a key observation is that
$$
T(\ell) = \inf\{t: V(t) > \ell\}, 
$$
and that 
$$
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell.
$$
By [Theorem 3.1, @MeerschaertStoev08], we have the stochastic process 
convergence 
$$
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t), \quad t > 0.
$$
where $Y(t)$ is a time-changed ("subordinated") extremal process, and where 
$\tilde b(c)$ is a regularly varying norming function which is
_inverse_ to $b(c)$, in the sense that 
$b(\tilde b(c)) \sim c \sim \tilde b(b(c))$ [@BinghamGoldieTeugels].

Without loss of generality, we choose $\ell^*$ such that $F(\ell^*) = 1/e$, 
and let $\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))$. 
We may then calculate 
$$
\mathbf P\left[ \frac{T(\ell)}{b(1/p_{\ell})} > t \right]
= \mathbf P[T(\ell) > b(1/p_{\ell}) t]
= \mathbf P[V(ct) \le \ell]
$$
where we have substituted $c = b(1/p_{\ell})$. Moreover 
$$
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right]
$$
Defining the hitting time of level $\ell^*$ by $Y(t)$ as 
$\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}$,
we then have 
$$
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
$$
by [Proposition 4.2, @MeerschaertStoev08], where $X$ is an exponential 
random variable with mean $1$. 
Using [Theorem 19.1, @Haubold11], we see that 
$X^{1/\beta} D \sim {\rm ML}(\beta, 1)$, concluding the proof. 
\qed

@@

1. Extract the $K$ largest order statistics 
(i.e. the $K$ largest values, where e.g. $K = 300$) 
$X_{(1)}, \ldots, X_{(K)}$ together with their timestamps $T_{(1)}, \ldots, T_{(K)}$.

2. Choose e.g. $K_0 = 5$, and for each $k$ ranging from $K_0 +1$ to $K + 1$: 
    a) extract the set $\mathcal T_k$ of exceedance times of the threshold $X_{(k)}$
    b) fit a Mittag-Leffler distribution to $\mathcal T_k$, resulting in the parameter
    estimates $\hat\beta_k$ and $\hat \sigma_k$.

3. Plot $k$ vs.\ $\hat \beta_k$. As $k \downarrow K_0$,
  $\hat \beta_k$ *stabilizes* around a constant $\hat \beta$.

4. Plot $k$ vs.\ $k^{1/\hat \beta} \hat \sigma_k$. 
  Where the plot stabilizes as $k \downarrow K_0$, define the value 
  $\hat \sigma_0$.

@@

grid.arrange(
  logMom_Ex1_tail,
  mle_Ex1_tail,
  hill_Ex1_tail,
  logMom_Ex2_tail,
  mle_Ex2_tail,
  hill_Ex2_tail,
  logMom_Ex3_tail,
  mle_Ex3_tail,
  hill_Ex3_tail,
  ncol = 3
)

```

@@

grid.arrange(
  logMom_Ex1_scale,
  mle_Ex1_scale,
  logMom_Ex2_scale,
  mle_Ex2_scale,
  logMom_Ex3_scale,
  mle_Ex3_scale,
  ncol = 2
)
```

>>>>

*Proof of Theorem:* We interpret the threshold crossing time $T(\ell)$ as the hitting time of the 
underlying CTRM (Continuous Time Random Maxima) or "max-renewal process", 
and then utilize a result by @MeerschaertStoev08. 
The running maximum process is defined as 
$$
M(c) := J_1 \vee \ldots \vee J_{\lfloor c \rfloor},
$$
and since we assume that the $J_k$ have a continuous distribution, there exist 
norming functions $a(c)$ and $d(c)$ such that 
$$
\mathbf P\left[ \frac{M(c) - d(c)}{a(c)} \le \ell^* \right] 
\longrightarrow F(\ell^*), \quad c \to \infty
$$
where $F$ is a generalized extreme value distribution, and $\ell^*$ is any 
value from the support of $F$. 
The CTRM process is then defined via 
$$
V(t) = M(N(t)), \quad t \ge 0
$$
where $N(t)$ is the renewal process associated with the waiting times 
$W_k$: 
$$
N(t) = \max\{n: W_1 + \ldots + W_n \le t\}.
$$
Now a key observation is that
$$
T(\ell) = \inf\{t: V(t) > \ell\}, 
$$
and that

\begin{align}
T(\ell) > t \quad \text{ if and only if } \quad V(t) \le \ell. \label{inverserelat}
\end{align}

By [Theorem 3.1, @MeerschaertStoev08], we have the stochastic process 
convergence

\begin{align}
\frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\stackrel{d}{\longrightarrow} Y(t),  \quad c \to \infty, \label{TheoMS}
\end{align}

for $t > 0$, where $Y(t)$ is a time-changed ("subordinated") extremal process, and where 
$\tilde b(c)$ is a regularly varying norming function which is
_inverse_ to $b(c)$, in the sense that 
$b(\tilde b(c)) \sim c \sim \tilde b(b(c))$ [@BinghamGoldieTeugels].

Without loss of generality, we choose $\ell^*$ such that $F(\ell^*) = 1/e$, 
and let $\ell = a(\tilde b(c)) \ell^* + d(\tilde b(c))$. 
We may then calculate 
$$
\mathbf P\left[ \frac{T(\ell)}{b(1/p_{\ell})} > t \right]
= \mathbf P[T(\ell) > b(1/p_{\ell}) t]
= \mathbf P[V(ct) \le \ell]
$$
where we have substituted $c = b(1/p_{\ell})$ and used \eqref{inverserelat}. Moreover, we receive due to the convergence in \eqref{TheoMS}
$$
\mathbf P[V(ct) \le \ell]
= \mathbf P\left[ \frac{V(ct) - d(\tilde b(c))}{a(\tilde b(c))} 
\le \frac{\ell - d(\tilde b(c))}{a(\tilde b(c))} \right]
\longrightarrow \mathbf P\left[ Y(t) \le \ell^* \right] ,  \quad c \to \infty,
$$
for $t>0$. Defining the hitting time of level $\ell^*$ by $Y(t)$ as 
$\xi_{\ell^*} := \inf\{t: Y(t) > \ell^*\}$,
we then have 
$$
P\left[ Y(t) \le \ell^* \right] = \mathbf P[\xi_{\ell^*} > t] 
= \mathbf P[(-\log F(\ell^*))^{-1/\beta} X^{1/\beta} D > t]
$$
by [Proposition 4.2, @MeerschaertStoev08], where $X$ is an exponential 
random variable with mean $1$. 
Using [Theorem 19.1, @Haubold11], we see that 
$X^{1/\beta} D \sim {\rm ML}(\beta, 1)$, concluding the proof. 
\qed

@@

1. Extract the $K$ largest order statistics 
(i.e. the $K$ largest values, where e.g. $K = 300$) 
$X_{(1)}, \ldots, X_{(K)}$ together with their timestamps $T_{(1)}, \ldots, T_{(K-1)}$.

2. Choose e.g. $K_0 = 5$, and for each $k$ ranging from $K_0 +1$ to $K + 1$: 
    a) extract the set $\mathcal T_k$ of exceedance times of between the magnitudes exceeding the threshold $X_{(k)}$
    b) fit a Mittag-Leffler distribution to $\mathcal T_k$, resulting in the parameter
    estimates $\hat\beta_k$ and $\hat \sigma_k$.

3. Plot $k$ vs.\ $\hat \beta_k$. To the right, as $k \downarrow K_0$, the asymptotics are off and bias is high. To the left (high threshold), data are scarce and variance is high. Choose a region of *stability* in the middle for a parameter estimate $\hat \beta$.

4. Plot $k$ vs.\ $k^{1/\hat \beta} \hat \sigma_k$. Again, choose a region of *stability* as $k$ gets smaller and define $\hat \sigma_0$.

@@

grid.arrange(
  arrangeGrob(logMom_Ex1_tail,logMom_Ex2_tail,logMom_Ex3_tail,top="log-Moment"),
  arrangeGrob(mle_Ex1_tail, mle_Ex2_tail, mle_Ex3_tail,top="MLE"),
  arrangeGrob(hill_Ex1_tail,hill_Ex2_tail,hill_Ex3_tail,top="Hill"),
  ncol=3
)
```

@@

grid.arrange(
  arrangeGrob(logMom_Ex1_scale,logMom_Ex2_scale,logMom_Ex3_scale,top="log-Mom"),
  arrangeGrob(mle_Ex1_scale,mle_Ex2_scale,mle_Ex3_scale,top="MLE"),
  ncol = 2
)
```
